---
title: "Survival Analysis with episodes and tidymodels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{survival}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{episodes, tidymodels, censored, survival, ggsurvfit, scales, dplyr}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The {episodes} package segments longitudinal data into meaningful
episodes based on temporal gaps and variable changes. It's designed for
analyzing treatment patterns, patient journeys, and other time series
data where interruptions and changes in covariates are important to
model.

This vignette demonstrates how to apply these features to survival
analysis using {tidymodels} to run Cox proportional hazards models with
both fixed and time-varying covariates. I will be using the simulated
`substance_use` dataset included in the package.

## Required Packages

```{r}
library(episodes)
library(tidymodels)
library(censored)
library(survival)
library(ggsurvfit)
library(dplyr)

surv_spec <- proportional_hazards() |>
  set_engine("survival")
```

## Time to Discontinuation Analysis

### Creating Episodes with Fixed Covariates

Our first example examines time to treatment discontinuation, using a
fixed-effects approach where covariates are considered constant
throughout the episode.

```{r}
episodes_fixed <- substance_use |>
  group_by(client_id) |>
  segment_episodes(
    date_col = visit_date,
    gap_threshold = 60,
    gap_unit = "days",
    progress = FALSE
  )

model_result <- workflow() |>
  add_formula(
    Surv(episode_days, discontinued) ~
      age + gender + race + substance + medication + mental_health_disorder +
      died_ever + cluster(client_id)
  ) |>
  add_model(surv_spec) |>
  fit(data = episodes_fixed) |>
  extract_fit_parsnip() |>
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.95)

model_result
```

#### Interpreting the Results

The model results show hazard ratios (HR) for each variable, where: - HR
\< 1: Reduced risk of discontinuation (better retention) - HR \> 1:
Increased risk of discontinuation (worse retention) - HR = 1: No effect
on discontinuation risk

Key findings:

-   **Mental health disorder**: Clients with mental health disorders
    have significantly better retention (HR = 0.689, 95% CI:
    0.657-0.722), indicating a 31.1% reduction in discontinuation risk.

-   **Mortality**: Clients who died have a 40% higher risk of treatment
    discontinuation (HR = 1.40, 95% CI: 1.32-1.47).

-   **Substance type**: Opiate users have a 15% higher risk of
    discontinuation compared to alcohol users (HR = 1.15, 95% CI:
    1.06-1.25).

-   **Demographics**: Age, gender, and race show minimal effects on
    treatment retention in this model.

-   **Medications**: None of the medications show statistically
    significant effects on retention, though point estimates suggest
    small protective effects for most treatments.

### Visualizing Survival by Mental Health Disorder Status

We can visualize these differences with Kaplan-Meier curves:

```{r}
surv_curves <- survfit(
  Surv(episode_days, discontinued) ~ mental_health_disorder,
  data = episodes_fixed
)

ggsurvfit(surv_curves) +
  add_confidence_interval() +
  add_risktable() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Days",
    y = "Probability of Retention",
    title = "Kaplan-Meier Curves by Mental Health Disorder Status"
  )
```

The Kaplan-Meier plot confirms our model findings - clients with mental
health disorders (blue line) have substantially better retention over
time compared to those without (red line). At 500 days, approximately
40% of clients with mental health disorders remain in treatment,
compared to only about 20% of those without. The non-overlapping
confidence intervals indicate this difference is statistically
significant throughout the observation period.

### Time-Varying Covariates Model

The `segment_episodes_by_covars` function creates a dataset where
observations are split whenever a specified covariate changes value.
This allows us to incorporate time-varying covariates in survival
analysis.

```{r}
episodes_tv <- substance_use |>
  group_by(client_id) |>
  segment_episodes_by_covars(
    date_col = visit_date,
    covar_cols = c(
      "frequency_of_use", "housing"
    ),
    gap_threshold = 60,
    gap_unit = "days",
    progress = FALSE
  ) |>
  group_by(client_id, episode_id) |>
  add_surv_time()

tv_model_result <- workflow() |>
  add_formula(
    Surv(time = start_time, time2 = end_time, event = discontinued) ~
      age + gender + race + substance + medication + mental_health_disorder +
      frequency_of_use + housing + died_ever + cluster(client_id)
  ) |>
  add_model(surv_spec) |>
  fit(data = episodes_tv) |>
  extract_fit_parsnip() |>
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.95)

tv_model_result
```

#### Interpreting Time-Varying Results

When we include time-varying covariates, we gain additional insights:

-   **Mental health disorder** continues to show a strong protective
    effect (HR = 0.678, 95% CI: 0.647-0.711), consistent with our
    fixed-effects model.
-   **Housing status** emerges as a significant predictor (housing.L, HR
    = 1.62, 95% CI: 1.53-1.72), suggesting clients with unstable housing
    have a 62% higher risk of discontinuation.
-   **Mortality risk** shows a slightly attenuated effect (HR = 1.21 vs.
    1.40 in the fixed model), suggesting some of the association is
    explained by time-varying factors.
-   **Substance type**: Opiate use remains associated with higher
    discontinuation risk (HR = 1.17, 95% CI: 1.07-1.27).
-   **Frequency of use** shows no clear linear relationship with
    retention.

The time-varying model provides a more nuanced understanding of
retention dynamics, highlighting the importance of housing stability as
a key factor in treatment continuation.

## Time to Death (Last Episode)

### Last Episode Analysis

We can also analyze time to death, focusing only on the last episode for
each client:

```{r}
episodes_fixed_death <- substance_use |>
  group_by(client_id) |>
  segment_episodes(
    date_col = visit_date,
    gap_threshold = 60,
    gap_unit = "days",
    episodes = "last",
    progress = FALSE
  )

death_model_result <- workflow() |>
  add_formula(
    Surv(episode_days, died_ever) ~
      age + gender + race + substance + medication + mental_health_disorder
  ) |>
  add_model(surv_spec) |>
  fit(data = episodes_fixed_death) |>
  extract_fit_parsnip() |>
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.95)

death_model_result
```

#### Interpreting Mortality Results

The mortality analysis reveals significant protective and risk factors:

-   **Mental health disorder**: Strong protective association with
    mortality (HR = 0.405, 95% CI: 0.363-0.452), suggesting a 59.5%
    reduction in mortality risk for clients with mental health
    disorders.
-   **Substance type**: Opiate users have a 29% higher mortality risk
    compared to alcohol users (HR = 1.29, 95% CI: 1.08-1.54).
-   **Race**: "Other" race category shows elevated mortality risk (HR =
    1.33, 95% CI: 1.05-1.70).
-   **Medications**: While not reaching statistical significance at
    p\<0.05, there's a trend toward protective effects for several
    medications, particularly Acamprosate (HR = 0.853) and Naltrexone
    (HR = 0.859).

These findings highlight the complex relationship between mental health
comorbidity and substance use outcomes, with mental health disorders
associated with both better retention and lower mortality.

### Visualizing Survival by Substance Type

```{r}
surv_curves_substance <- survfit(
  Surv(episode_days, died_ever) ~ substance,
  data = episodes_fixed_death
)

ggsurvfit(surv_curves_substance) +
  add_confidence_interval() +
  add_risktable() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Days",
    y = "Probability of Survival",
    title = "Kaplan-Meier Curves by Substance"
  )
```

The survival curves by substance type show that opiate users (red line)
have consistently worse survival outcomes compared to alcohol users
(blue line). The separation of curves begins early and widens over time,
with approximately 85% of alcohol users surviving to 1000 days compared
to only about 75% of opiate users. The risk table shows how the number
at risk decreases over time for both groups.

### Time-Varying Death Analysis

For a more nuanced analysis, we can incorporate time-varying covariates:

```{r}
episodes_tv_death <- substance_use |>
  group_by(client_id) |>
  segment_episodes_by_covars(
    date_col = visit_date,
    covar_cols = c(
      "died_last_visit", "frequency_of_use", "housing"
    ),
    gap_threshold = 60,
    gap_unit = "days",
    progress = FALSE
  ) |>
  group_by(client_id, episode_id) |>
  add_surv_time()

death_tv_model <- workflow() |>
  add_formula(
    Surv(time = start_time, time2 = end_time, event = died_last_visit) ~
      age + gender + race + substance + medication + mental_health_disorder +
      frequency_of_use + housing + cluster(client_id)
  ) |>
  add_model(surv_spec) |>
  fit(data = episodes_tv_death) |>
  extract_fit_parsnip() |>
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.95)

death_tv_model
```

#### Interpreting Time-Varying Mortality Results

The time-varying mortality model reveals additional important factors:

-   **Housing status** emerges as the strongest predictor (housing.L, HR
    = 3.25, 95% CI: 2.91-3.62), indicating that unstable housing is
    associated with more than three times the risk of mortality.
-   **Substance type**: Opiate use shows an even stronger association
    with mortality (HR = 2.11, 95% CI: 1.76-2.53) than in the fixed
    model.
-   **Mental health disorder** remains strongly protective (HR = 0.505,
    95% CI: 0.452-0.565).
-   **Frequency of use** shows a complex non-linear relationship with
    mortality (both linear and quadratic terms significant).
-   **Naltrexone** emerges as significantly protective (HR = 0.812, 95%
    CI: 0.681-0.968), suggesting a 18.8% reduction in mortality risk.
-   **Age** becomes marginally significant (HR = 1.00, p = 0.039),
    suggesting a small increase in mortality risk with age.

The time-varying model highlights how critical social determinants of
health (housing) and current use patterns are in predicting mortality
risk beyond static client characteristics.

## Time to Death (All Episodes)

### Basic Model with Clustering

When analyzing data across all episodes for each client, we use
clustering to account for within-client correlation:

```{r}
episodes_clustered <- substance_use |>
  group_by(client_id) |>
  segment_episodes_by_covars(
    date_col = visit_date,
    covar_cols = c("died_last_visit"),
    gap_threshold = 60,
    gap_unit = "days",
    progress = FALSE
  ) |>
  group_by(client_id, episode_id) |>
  add_surv_time()

clustered_model <- workflow() |>
  add_formula(
    Surv(time = start_time, time2 = end_time, event = died_last_visit) ~
      age + gender + race + substance + medication + mental_health_disorder +
      cluster(client_id)
  ) |>
  add_model(surv_spec) |>
  fit(data = episodes_clustered) |>
  extract_fit_parsnip() |>
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.95)

clustered_model
```

#### Interpreting Clustered Results

When analyzing mortality across all episodes with clustering:

-   **Mental health disorder** maintains its protective association (HR
    = 0.471, 95% CI: 0.421-0.526).
-   **Substance type**: Opiate use shows a strong association with
    mortality (HR = 2.21, 95% CI: 1.84-2.64).
-   **Medications**: Both Acamprosate (HR = 0.811, 95% CI: 0.666-0.988)
    and Naltrexone (HR = 0.823, 95% CI: 0.691-0.980) show statistically
    significant protective effects.
-   **Race**: The "Other" race category continues to show elevated risk
    (HR = 1.33, 95% CI: 1.04-1.69).

By accounting for within-client correlation using the `cluster()`
function, we ensure that standard errors are appropriately adjusted for
repeated observations from the same clients.

### Comprehensive Model

Finally, we build a more comprehensive model with all available
time-varying covariates:

```{r}
episodes_comprehensive <- substance_use |>
  group_by(client_id) |>
  segment_episodes_by_covars(
    date_col = visit_date,
    covar_cols = c(
      "died_last_visit", "discontinued", "frequency_of_use", "housing"
    ),
    gap_threshold = 60,
    gap_unit = "days",
    progress = FALSE
  ) |>
  group_by(client_id, episode_id) |>
  add_surv_time()

comprehensive_model <- workflow() |>
  add_formula(
    Surv(time = start_time, time2 = end_time, event = died_last_visit) ~
      age + gender + race + substance + medication + mental_health_disorder +
      frequency_of_use + housing + discontinued + cluster(client_id)
  ) |>
  add_model(surv_spec) |>
  fit(data = episodes_comprehensive) |>
  extract_fit_parsnip() |>
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.95)

comprehensive_model
```

#### Interpreting the Comprehensive Model

The fully specified model reveals the most complete picture of mortality
risk factors:

-   **Treatment discontinuation** is a major risk factor (HR = 2.98, 95%
    CI: 2.66-3.34), with clients who discontinue treatment having nearly
    three times the mortality risk.
-   **Housing instability** remains a strong predictor (housing.L, HR =
    2.77, 95% CI: 2.48-3.09), even after controlling for other factors.
-   **Substance type**: Opiate use continues to show elevated risk (HR =
    2.22, 95% CI: 1.85-2.66).
-   **Mental health disorder** maintains its protective association (HR
    = 0.573, 95% CI: 0.513-0.641), though the effect size is somewhat
    attenuated in this full model.
-   **Frequency of use** shows both linear (HR = 1.18, 95% CI:
    1.02-1.37) and quadratic (HR = 1.34, 95% CI: 1.16-1.55)
    relationships with mortality.
-   **Naltrexone** continues to show a protective effect (HR = 0.822,
    95% CI: 0.690-0.980).

This comprehensive model highlights the critical interplay between
treatment retention, social determinants (housing), substance type, and
medication in determining mortality outcomes. The finding that treatment
discontinuation is associated with nearly three times the mortality risk
underscores the life-saving potential of effective retention strategies.

## Limitations

Keep in mind that these are simple Cox models to inspire users of
{episodes} to see how easy it is to get started running survival
analyses. Additionally, we are not testing any of the assumptions,
mainly the proportional hazards assumption.

Some limitations to consider:

1.  The proportional hazards assumption should be verified using tests
    like Schoenfeld residuals (`cox.zph()` in the {survival} package)

2.  Recurrent events are treated independently, ignoring temporal
    ordering effects

3.  Competing risks and multiple states (discontinuation and death) are
    not accounted for, which may lead to biased estimates

## R Package Suggestions

If you're interested in exploring more complex modeling approaches, you
could consider:

### For Cox Models

-   {survival}: The foundation for survival analysis in R with functions
    for Cox regression, Kaplan-Meier curves, and diagnostics

-   {survminer}: Enhanced visualization of survival analyses

-   {coxme}: Mixed-effects Cox models for clustered/grouped data

### More Complex Models

-   {cmprsk} or {riskRegression}: Competing risks analysis

-   {frailtypack}: Joint frailty models for recurrent events and terminal events

-   {msm}: Multi-state models for transitions between states

-   {mstate}: Multi-state modeling with the Cox proportional hazards model

### Assumptions Testing

-   {flexsurv}: Flexible parametric survival models

-   {timereg}: For non-proportional hazards models

### Bayesian Approaches

-   {rstanarm}: Bayesian survival models using Stan

-   {brms}: Flexible Bayesian regression models including survival models

## Conclusion

The {episodes} package provides powerful tools for analyzing
longitudinal health data:

1.  `segment_episodes()` creates episode-level datasets based on
    temporal gaps
2.  `segment_episodes_by_covars()` allows for time-varying covariate
    analysis
3.  `add_surv_time()` prepares data for survival analysis

These functions integrate seamlessly with {tidymodels} and survival
analysis workflows, enabling sophisticated analyses of treatment
patterns, outcomes, and risk factors.

By segmenting longitudinal data into meaningful episodes and
incorporating time-varying covariates, researchers can gain deeper
insights into the complex dynamics of substance use treatment outcomes.
